Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan

### 2020-01-04 analysis/paper/paper.Rmd
move a paragraph to scrape file

To identify craft specialization, the variation of standardization of specialized product is commonly used as indicator based on the assumption that specialized mass production will lead to uniformity of the product due to routinization, increased skills, and fewer number of producers involved in production [@Stark1995; @Costin1991; @Costin2001]. For ceramic standardization, several measurements have been proposed, such as metric, compositional, and technological approaches [@Arnold2000; @Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015]. Among those variables, metric measurement is widely applied in archaeological assemblages by calculating coefficient of variation between sets of data coupled with statistical comparison that lower value reflects higher degree of standardization [@Eerkens2001; @Roux2003; @Roux2018; @Stark1995]. However, this traditional typological and linear measurements might be limited because they can be insensitive to subtle variations resulting from changes in assembles. Shape analysis based on geometric morphometrics method are new approaches for measuring the degree of standardization since it takes the overall shape as variable for comparison [@Slice2007]. Taking geometric morphomerics approaches to the measurement of craft specialization, we study shape and standardization of locally made ceramic at Kiwulan (1350-1950 AD) [@Chen2007], a large multi-component archaeological site in northeastern Taiwan, to identify changes in ceramic production in cross-cultural interaction contexts. The aim of this paper is to investigate if there are increasing ceramic specialization resulting from interaction with the Europeans and the Chinese, two major foreign influences in early historical Taiwan that might indicate social changes in the indigenous society.

Our hypothesis is that if foreign contacts including European in the 17th century and Chinese in the 19th century had impacts on the emergence of social inequality in local indigenous society due to the monopolies for trade between a small number of indigenous people, then we expect to see the changes in social organization from a more corporate to a more network organization that would be reflected by pottery production at Kiwulan. The emergence of craft specialization, here pottery specialization, are usually related to the presence of elite control. In this case, if the competitions for foreign resources and being trade partners of European or Chinese colonizer among individuals gradually lead to the emergence of social inequality, then we expect the local pottery will show more homogeneous features after contact due to the craft specialization caused by control of small group of individuals. The aim of this paper is to explore the impact of foreign contacts on local indigenous societies in northeastern Taiwan by examining pottery production as an indicator for social changes. Using pottery shape as a proxy to study craft specialization, we answer these questions: Did foreign contacts have impacts on indigenous pottery production that can be detected in the shape of the vessels? How does pottery change in shape throughout 500 years at this site? Did pottery become more homogeneous and standardized in shape after foreign contacts with the European colonizer or the Chinese immigrant?

## Geometric Morphometric approach: shape analysis for ceramic specialization

@@

In this paper, we use EFA for pottery shape analysis to evaluate the level of standardization of pottery from an Iron Age site in northeastern Taiwan in relation to the European presence in the 17th century that might hint the emergence of craft specialization for pottery production. Craft specialization can provide further evidence about the type of production organization which is closely related to economic or political aspects of society [@Arnold2000; @Blackman1993; @Costin1991; @Hirshman2010; @Stark1995]. The emergence of craft specialization can also be an indicator of social inequality that hints the possibility of the presence of production groups worked for specific individuals. The presence of craft specialization is commonly studied by measuring the standardization of ceramics based on the assumption that the specialized mass production will lead to uniformity or homogeneity of the product due to increased skills, routinization, and lower diversity of producers [@Arnold2000]. To test the prediction of a shift in the level of social inequality in this case study in northeastern Taiwan influenced by the colonial European presence, we use standardization of pottery in shapes as an index for craft specialization to identify the presence of a few individuals controlling large scale pottery production [@Costin1995; @Hirshman2010] using outline-based geometric morphometrics couples with traditional metric measurements to compare and evaluate the use of outline approaches for pottery with high consistency in shapes.

- craft or pottery specialization
- CV is hard to test, we apply some statistics testing for it, - innovative: CV test

>>>>

To identify craft specialization, the variation of standardization of specialized product is commonly used as indicator based on the assumption that specialized mass production will lead to uniformity of the product due to routinization, increased skills, and fewer number of producers involved in production [@Stark1995; @Costin1991; @Costin2001]. For ceramic standardization, several measurements have been proposed, such as metric, compositional, and technological approaches [@Arnold2000; @Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015]. Among those variables, metric measurement is widely applied in archaeological assemblages by calculating coefficient of variation between sets of data coupled with statistical comparison that lower value reflects higher degree of standardization [@Eerkens2001; @Roux2003; @Roux2018; @Stark1995]. However, this traditional typological and linear measurements might be limited because they can be insensitive to subtle variations resulting from changes in assembles. Shape analysis based on geometric morphometrics method are new approaches for measuring the degree of standardization since it takes the overall shape as variable for comparison [@Slice2007]. Taking geometric morphomerics approaches to the measurement of craft specialization, we study shape and standardization of locally made ceramic at Kiwulan (1350-1950 AD) [@Chen2007], a large multi-component archaeological site in northeastern Taiwan, to identify changes in ceramic production in cross-cultural interaction contexts. The aim of this paper is to investigate if there are increasing ceramic specialization resulting from interaction with the Europeans and the Chinese, two major foreign influences in early historical Taiwan, that might indicate social changes in the indigenous society.

Our hypothesis is that if foreign contacts including European in the 17th century and Chinese in the 19th century had impacts on the emergence of social inequality in local indigenous society due to the monopolies for trade between a small number of indigenous people, then we expect to see the changes in social organization from a more corporate to a more network organization that would be reflected by pottery production at Kiwulan [@Feinman2000]. The emergence of craft specialization, here pottery specialization, are usually related to changes in social organization towards a society with increasing status inequality. In this case, if the competitions for foreign resources and being trade partners of European or Chinese colonizer among individuals gradually lead to the emergence of social inequality, then we expect the local ceramic will show more homogeneous features after contact due to the craft specialization caused by control of small group of individuals. Using pottery shape as a proxy to study craft specialization, we answer these questions: Did foreign contacts have impacts on indigenous pottery production that can be detected in the shape of the vessels? How does pottery change in shape throughout 500 years at this site? Did pottery become more homogeneous and standardized in shape after foreign contacts with the European colonizer or the Chinese immigrant? Those question are important to understand the indirect influences of the European colonists on the local indigenous societies. Also, it would help to understand the relationship between craft specialization and increasing social inequality in a pericolonial context [@Acabado2017].

## Geometric Morphometric approaches

@@

In this paper, we use EFA for pottery shape analysis to evaluate the level of standardization of pottery from an Iron Age site in northeastern Taiwan in relation to the European presence in the 17th century that might hint the emergence of craft specialization for pottery production.

- CV is hard to test, we apply some statistics testing for it, - innovative: CV test
